---
title: "Covid Mutation rates (Bayesian approach)"
author: "Felix Barenys Marimon"
date: "1/8/2022"
output: 
 pdf_document:
    toc: false
    number_sections: true
---

\pagebreak
```{r include=FALSE}
library(knitr)
```


\clearpage

# Abstract 

The aim of this study is to analyze the mutation rate of the COVID-19 virus based on the data available in the cov-lineages^[https://cov-lineages.org/lineage_list.html] website and use Bayesian statistical methods in order to be able to estimate the real mutation rate of the virus based on an article^[https://www.bath.ac.uk/announcements/mutation-rate-of-covid19-virus-is-at-least-50-per-cent-higher-than-previously-thought/] that estimates that the real mutation rate is 40% higher than the one seen in data.
\clearpage

\tableofcontents
\pagebreak

# Data Handling

In this article we will work with the time difference (in days ) between days recorded in the website.

```{r message=FALSE, warning=FALSE}
covid_rep<-read.csv("/Users/felixbarenysmarimon/Desktop/PROJECT/mutations covid/covid_variants.csv")
covid_rep<-covid_rep[,c(-1,-2,-4,-6,-8,-10,-12,-14,-15,-16)]
covid_rep$Earliest.date<-as.Date(covid_rep$Earliest.date)
covid_rep<-covid_rep[order(covid_rep$Earliest.date),]

#establishing which are the worrying mutations
worring<-c("B.1.351 ","P.1 ","B.1.617.2 ","B.1.1.529 ")
covid_rep$worry<-"No"
covid_rep$worry[which(is.element(covid_rep$Lineage,worring)==TRUE)]<-c("Beta",
                                                          "Gamma","Delta","Omicron")

#establish alpha for plot color

worring<-c("B.1.351 ","P.1 ","B.1.617.2 ","B.1.1.529 ")
covid_rep$alpha<-0.1
covid_rep$alpha[which(is.element(covid_rep$Lineage,worring)==TRUE)]<-1


#Eliminate Missings:
covid_rep<-covid_rep[-which(is.na(covid_rep$Earliest.date)),]


#difference between mutations in time(days)
covid_rep$diff_time<-c(0,as.numeric(diff.Date(covid_rep$Earliest.date)))
cov<-covid_rep$diff_time
```

# Data distribution

Here the distribution of the data is shown below:

```{r}
tab<-t(table(cov))
kable(tab,caption="Days between mutations/ nº mutations")
```
If we consider that the mutations occur independently in time, the time between mutations should follow a poisson with lambda `r round(mean(cov),3)`:


```{r}
hist(covid_rep$diff_time,xlim=c(0,12),breaks=60,
     xlab="Difference between mutations",main="Histogram",
     probability =TRUE)
lines(0:max(cov), dpois(0:max(cov), mean(cov)), col = 'red')
```


# Bayesian model

First we will compute the likehood function of the Poisson model:

$$L(\lambda;x_1...x_n)=\prod_{i=1}^nexp(-\lambda)\frac{1}{x_i!}\lambda^{x_i} = exp(-n\lambda)\frac{\lambda^{\sum_{i=1}^n x_i}}{\prod_{i=1}^ni=n!}$$ and with our data:

$$L(\lambda)=exp(-1536\lambda)\frac{\lambda^{702}}{4.274883e+69} \propto exp(-1536\lambda)\lambda^{702}$$

A good prior distribution from this data could be a Gamma because of its proprieties, which could be centered  `r round(mean(cov),4)` which is a 50% higher than the lambda of the distirbution of the data and a with a standard deviation of 0.01:
Now we can estimate the parameters of the gamma distribution having in account that:

$$\pi(\lambda)=\frac{\beta^{\alpha}( \lambda)^{\alpha-1}e^{-\beta \lambda}}{\Gamma(\alpha)}$$ with:

$\alpha=\frac{\bar{x}^2}{s^2}$ and $\beta=\frac{\bar{x}}{s^2}$ 

```{r}
alpha=(1.5*mean(cov))^2/0.01
beta=(1.5*mean(cov))/0.01

curve(dgamma(x,alpha,beta))
```

And then we can calculate the posterior having in account that a posterior from a Poisson and Gamma distributions in equal to:

$$f(\lambda/X) \propto L(X/\lambda)\pi(\lambda) \propto \lambda^{\alpha+ \sum^n_{i=1}x_i-1}e^{-(\beta+n)\lambda}$$ that in this case is:
$$f(\lambda/X) \propto  \lambda^{747}e^{-1604.5\lambda}$$ 
that is proportional to a  Gamma(748,1604.55).
\linebreak

```{r}
curve(dgamma(x,alpha+sum(cov),beta+length(cov)),ylab="Posterior distribution",xlab="Lambda")

```



# Credibility intervals

With the posterior lambda distribution we can calculate a credibility interval for the lambda:

```{r}
lower<-qgamma(0.025,alpha+sum(cov),beta+length(cov));lower
upper<-qgamma(0.975,alpha+sum(cov),beta+length(cov));upper

```

And then we can plot the distribution with the credibility intervals:

```{r}
hist(covid_rep$diff_time,xlim=c(0,12),breaks=60,
     xlab="Difference between mutations",main="Histogram",
     probability =TRUE)
lines(0:max(cov), dpois(0:max(cov), mean(cov)), col = 'red')
lines(0:max(cov), dpois(0:max(cov), lower), col = 'purple')
lines(0:max(cov), dpois(0:max(cov), upper), col = 'green')
```

\pagebreak

# Predictive posterior distirbution

The predictive posterior distribution is defined as:

$$p(X_{new}/X)=\int_{\Theta}p(X_{new}/\theta)f(\theta/X) d\theta$$ being $\Theta$ the domain  of the parameter. With our posterior and the Poisson distribution from the data we have:

$$p(X_{new}/X)=\int_{0}^{\infty}e^{-\lambda} \frac{1}{x_{new}!}\lambda^{x_{new}}\frac{1604.5^{748}( \lambda)^{747}e^{-1604.5 \lambda}}{\Gamma(748)} d\lambda $$

$$= \frac{1604.5^{748}*\Gamma(x_{new}+748)}{x_{new}!\Gamma(748)1605.5^{x_{new}+748}}=\frac{\Gamma(x_{new}+748)}{x_{new}!\Gamma(748)1605.5^{x_{new}}}$$ and having in account that $\Gamma(748)\simeq \infty$,
$$ p(X_{new}/X)=\frac{1}{x_{new}!1605.5^{x_{new}}} $$

```{r}
predpost<-function(x){1/(factorial(x)*1605^x)}
curve(predpost(x))
```



# Conclusions

Applying the Bayesian approach to the data, the real mutation rate (considering the Gamma prior), should be between `r lower` and `r upper`.

More in depth exploration and analysis must be done in order to be able to predict more accurately the mutation nature of the virus.
